---
title: "Binary Treatment: Average Treatment Effect"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary Treatment: Average Treatment Effect}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

This vignette demonstrates frengression for binary treatment data.
We estimate the Average Treatment Effect (ATE) and compare with a
simple doubly-robust estimator.

## Data Generating Process

- $Z \sim N(0, 1)$
- $X \sim \text{Bernoulli}(\text{logit}^{-1}(Z))$ (confounded treatment)
- $Y = 2X + Z + \varepsilon$, where $\varepsilon \sim N(0, 0.5^2)$

True ATE = $E[Y|do(X=1)] - E[Y|do(X=0)] = 2$.

```{r data}
library(Rfrengression)
set.seed(123)

n <- 3000
z <- matrix(rnorm(n), ncol = 1)
prob_x <- plogis(z)
x <- matrix(rbinom(n, 1, prob_x), ncol = 1)
y <- 2 * x + z + matrix(rnorm(n, sd = 0.5), ncol = 1)

cat("Treatment prevalence:", mean(x), "\n")
cat("Naive difference-in-means:", mean(y[x == 1]) - mean(y[x == 0]), "\n")
cat("(biased due to confounding)\n")
```

## Train Frengression

```{r train, eval = torch::torch_is_installed()}
model <- frengression(x_dim = 1, y_dim = 1, z_dim = 1,
                      x_binary = TRUE, noise_dim = 5)

model <- train_y(model, x, z, y, num_iters = 500, lr = 1e-3,
                 print_every = 100)
model <- train_xz(model, x, z, num_iters = 500, lr = 1e-4,
                   print_every = 100)
```

## Estimate ATE

```{r ate, eval = torch::torch_is_installed()}
# E[Y | do(X=1)]
y1 <- predict(model, matrix(1, ncol = 1), type = "mean", nsample = 1000)
# E[Y | do(X=0)]
y0 <- predict(model, matrix(0, ncol = 1), type = "mean", nsample = 1000)

ate_freng <- y1 - y0
cat("Frengression ATE estimate:", ate_freng, "\n")
cat("True ATE: 2\n")
```

## Compare: Simple DR Estimator

```{r dr, eval = torch::torch_is_installed()}
# Simple DR estimator for comparison
# Convert matrix columns to vectors for lm/glm
xv <- as.numeric(x)
yv <- as.numeric(y)
zv <- as.numeric(z)

e_hat <- glm(xv ~ zv, family = binomial)$fitted.values
mu1_hat <- predict(lm(yv ~ zv, subset = xv == 1), newdata = data.frame(zv = zv))
mu0_hat <- predict(lm(yv ~ zv, subset = xv == 0), newdata = data.frame(zv = zv))

dr_ate <- mean(mu1_hat - mu0_hat +
               xv * (yv - mu1_hat) / e_hat -
               (1 - xv) * (yv - mu0_hat) / (1 - e_hat))
cat("Doubly-robust ATE estimate:", dr_ate, "\n")
```

## Visualize Causal Margin Distributions

```{r plot, eval = torch::torch_is_installed(), fig.cap = "Distribution of Y under do(X=0) and do(X=1)"}
y_samples_1 <- sample_causal_margin(model, matrix(1, ncol = 1), sample_size = 500)
y_samples_0 <- sample_causal_margin(model, matrix(0, ncol = 1), sample_size = 500)

hist(as.numeric(y_samples_1), breaks = 30, col = rgb(1, 0, 0, 0.4),
     main = "Causal Margin: P(Y | do(X))", xlab = "Y",
     xlim = range(c(as.numeric(y_samples_0), as.numeric(y_samples_1))))
hist(as.numeric(y_samples_0), breaks = 30, col = rgb(0, 0, 1, 0.4), add = TRUE)
legend("topright", c("do(X=1)", "do(X=0)"),
       fill = c(rgb(1, 0, 0, 0.4), rgb(0, 0, 1, 0.4)))
```
